---
title: "GVPT 2020 Fall Method Workshop: Modelling"
author: "Xiaonan Wang"
date: "November 17, 2020"
output:
  ioslides_presentation: default
  slidy_presentation: default
---

## 2020 GVPT Method WorKshop

-- **Manage work flows**

-- **Data collection**

-- **Data cleaning**

-- **Learn from your data (today)**

-- **Visualization: communicate to others (next workshop)**

## What Is Not Covered in This Presentation 

-- **Causal inference (next semester)**

-- **Basics and extensions of models (GVPT 622, 722, 729, ...)**

-- **Integrating theoretical and empirical models (EITM)**

-- **Machine learning**

## What This Presentation Is About

-- **How Models Can Benefit and Mislead Us**

  - Data generating process
  
  - Data versus model structure
  
  - Use intuitive examples to highlight practical trade-offs

-- **Use Modelling Tools from *Tidyverse* Package**

  - Tools that make data exploratory easier (know your data)

  - Work with pipes and seamlessly connect with data
  
  - *R for Data Science*, https://r4ds.had.co.nz/

## Data Generating Process

Let's say we have the following dataset

```{r,warning=FALSE,message=FALSE,echo=FALSE}
### Require tidyverse package
library(tidyverse)
### Set the seed of R's random number generator (for reproducibility) 
set.seed(1221)
### Number of observations
obs<-500
### Generate D
D<-sample(c(0,1),obs,replace=T)
### Generate X
mu_x<-1; sigma_x<-5
X<-rnorm(n=obs,mean=mu_x,sd=sigma_x)
### Generate Y (Y=b0+b1*D+b2*X+e)
b0<-0.5; b1<-1.6; b2<-1.8; mu_e<-0; sigma_e<-5 
y<-b0+b1*D+b2*X+rnorm(n=obs,mean=mu_e,sd=sigma_e)
### tibble y, D, and X as a dataset
d<-tibble(y,D,X)
d
```

## Data Generating Process

-- How are y, D, and X generated?

-- What distributions do y, D, and X come from?

-- Are observations correlated and how?

-- Are variables correlated and how?

......

**We often don't know the true data generating process**

## Model

-- **Models partition variations into patterns and noises**

  - Variation: within a variable
  
  - Covariation: between variables
  
  - A model is a low-dimension summary of the dataset

-- **Models are subjective**

  - "All models are wrong, but some are useful"

  - "A model is a representation for a particular purpose"

## Model and Data

-- **Models put structures on data**

  - Assumptions about connections among variables are made
  
  - Functional forms
  
  $$y=\alpha_0+\alpha_1 x_1+ \alpha_2 x_2 +\alpha_3 x_3$$

## Model and Data

-- **Why not just let data speak for itself**

  - There are ways to extract connections without structures
  
  - But structures may exist in the data
  
  - Easier to interpret
  
  - Easier to quantify uncertainty
  
  - We have theories (know the data generating process)
  
-- **We do not always have the luxury to let data speak for itself**  
  
  - **We need large and informative data!**
  
  - **Research designs!** (next semester) 

## Example 1

```{r,warning=FALSE,message=FALSE}
library(tidyverse)
### Set the seed of R's random number generator (for reproducibility) 
set.seed(1221)
### Number of observations
obs<-500
### Generate D
D<-sample(c(0,1),obs,replace=T)
### Generate X
mu_x<-1; sigma_x<-5
X<-rnorm(n=obs,mean=mu_x,sd=sigma_x)
### Generate Y (Y=b0+b1*D+b2*X+e)
b0<-0.5; b1<-1.6; b2<-1.8; mu_e<-0; sigma_e<-5  
y<-b0+b1*D+b2*X+rnorm(n=obs,mean=mu_e,sd=sigma_e)
```

**What does this data generating process tell us?**

## Example 1

```{r,warning=FALSE,message=FALSE}
### tibble y, D, and X as a dataset
d<-tibble(y,D,X)
### take a look at the data
summary(d)
```

## Example 1

**Research question: how does D affect Y?**

```{r,warning=FALSE,message=FALSE}
### compare two group means: Student-t test
t.test(y~D,var.equal=T)
```

## Example 1

```{r,warning=FALSE,message=FALSE}
### compare two group means (an equivalent way but using "lm")
gpmeans<-lm(data=d,y~D);summary(gpmeans)$coefficients
### adding x2, assuming a linear relationship y=beta0+beta1*D+beta2*X
mod<-lm(data=d,y~D+X);summary(mod)$coefficients
```


## Example 1

**Bias** (repeated sampling)

```{r,warning=FALSE,message=FALSE}
### a function to generate data
func_dgp<-function(N,mu_x,sigma_x,mu_e,sigma_e,b0,b1,b2){
  tibble(D=sample(c(0,1),N,replace=T),
         X=rnorm(N,mu_x,sigma_x),
         y=b0+b1*D+b2*X+rnorm(N,mu_e,sigma_e))
}
### a function to compare two group means
func_gpmeans<-function(dataset){
  m<-lm(data=dataset,y~D);coef(m)[2]
}
### a function to add X as a linear model
func_lmod<-function(dataset){
  m<-lm(data=dataset,y~D+X);coef(m)[2]
}
```

## Example 1

**Bias** (repeated sampling)

```{r,message=FALSE,warning=FALSE}
### repeated sampling and sampling distribution 
ndraws<-1000
est_gpmeans<-est_lmod<-list()
for (i in 1:ndraws){
  # draw 1000 samples
  d_repeated<-func_dgp(N=obs,mu_x=mu_x,mu_e=mu_e,
                       sigma_x=sigma_x,sigma_e=sigma_e,
                       b0=b0,b1=b1,b2=b2)
  # for each sample, compare the two group means
  est_gpmeans[[i]]<-func_gpmeans(data=d_repeated)
  # for each sample, run a linear model y~D+X
  est_lmod[[i]]<-func_lmod(data=d_repeated)
}
### tibble the sampling distributions
est_sampling<-tibble(est_1=unlist(est_gpmeans),est_2=unlist(est_lmod))
```


## Example 1

**Bias** (repeated sampling)

```{r,message=FALSE,warning=FALSE,echo=FALSE}
### plot the sampling distributions
ggplot(est_sampling)+
  geom_density(aes(est_1,fill="y~D"),alpha=0.2,lwd=1)+
  geom_density(aes(est_2,fill="y~D+X"),alpha=0.2,lwd=1)+
  scale_fill_manual(values=c("red","blue"),guide_legend(title=""))+
  geom_vline(xintercept = b1)+
  labs(x="sampling distributions")+
  theme_bw()
```

## Example 1

**Consistency** (number of observations grows to infinity)

```{r,message=FALSE,warning=FALSE}
sample_size<-seq(from=500,to=50000,by=100)
est_gpmeans<-est_lmod<-list()
for (i in 1:length(sample_size)){
  # draw samples with different sample_size
  d_asymptotic<-func_dgp(N=sample_size[i],mu_x=mu_x,mu_e=mu_e,
                       sigma_x=sigma_x,sigma_e=sigma_e,
                       b0=b0,b1=b1,b2=b2)
  # for each sample, compare the two group means
  est_gpmeans[[i]]<-func_gpmeans(data=d_asymptotic)
  # for each sample, run a linear model y~D+X
  est_lmod[[i]]<-func_lmod(data=d_asymptotic)
}
### tibble the sampling distributions
est_asymptotic<-tibble(est_1=unlist(est_gpmeans),
                       est_2=unlist(est_lmod))
```

## Example 1

**Consistency** (number of observations grows to infinity)

```{r,message=FALSE,warning=FALSE,echo=FALSE}
### plot the sampling distributions
ggplot(est_asymptotic)+
  geom_line(aes(y=est_1,x=1:length(sample_size),col="y~D"),lwd=1,alpha=0.6)+
  geom_line(aes(y=est_2,x=1:length(sample_size),col="y~D+X"),lwd=1,alpha=0.6)+
  scale_colour_manual(values=c("red","blue"),guide_legend(title=""))+
  geom_hline(yintercept = b1)+
  labs(x="number of observations",y="point estimates of b1")+
  scale_x_continuous(labels = sample_size[seq(1,length(sample_size),49)],
                     breaks=seq(1,length(sample_size),49))+
  theme_bw()
```

## Take-away Points from Example 1

**In a randomized experiment (or a very good research design)**

  -- Modelling is not essential in estimating the treatment effect
  
  -- Correctly specified models reduce standard errors
  
  -- Sample size matters
  
  -- With samller sampe sizes, correct model helps 
  
## Example 2

```{r,warning=FALSE,message=FALSE}
set.seed(1221)
### Number of observations
obs<-500
### Generate age
Age<-round(runif(n=obs,min=18,max=50))
### Generate gender
Gender<-sample(c(0,1),obs,replace=T)
### Generate D
xb<-(-9.5)+0.6*Age-0.01*I(Age^2)+2.8*Gender
p<-exp(xb)/(1+exp(xb))
D<-rbinom(n=obs,size=1,prob=p)
### Generate Y (Y=b0+b1*D+b2*Age+b3*Age^2+b4*Gender+e)
b0<-0.5; b1<-1.6; b2<-3.5; b3<-(-0.1); b4<-1.2
y<-b0+b1*D+b2*Age+b3*I(Age^2)+b4*Gender+
  rnorm(n=obs,mean=mu_e,sd=sigma_e)
```

**What does this data generating process tell us?**

## Example 2

```{r,warning=FALSE,message=FALSE}
### tibble y, D, and X as a dataset
d<-tibble(y,D,Age,Gender)
### take a look at the data
summary(d)
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### compare two group means
gpmeans<-lm(data=d,y~D)
summary(gpmeans)$coefficients
```

This is way off and we "know" why.

## Example 2

**Control covariates by putting structures on data: modelling**

$$y=\alpha_0+\alpha_1 D + \alpha_2 Age +\alpha_3 Gender$$

-- The effect of $D$ while holding $Age$ and $Gender$ equal

-- The assumption: this is the correct functional form

## Example 2

```{r,warning=FALSE,message=FALSE}
### modelling with the above functional form
mod<-lm(data=d,y~D+Age+Gender)
summary(mod)$coefficients
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### modelling with the correct functional form
mod_true<-lm(data=d,y~D+Age+I(Age^2)+Gender)
summary(mod_true)$coefficients
```

We often don't know this

## Example 2

**An intuitive way to control covariates: stratification**

--**compare cases in two groups with the same gender**

  -[D=1, Gender=1], [D=0, Gender=1]
  
  -[D=1, Gender=0], [D=0, Gender=0]
  
--**compare cases in two groups with the same gender and age**
  
  -[D=1, Gender=1, Age=25], [D=0, Gender=1, Age=25]
  
  -[D=1, Gender=0, Age=26], [D=0, Gender=0, Age=26]

## Example 2
  
-- **Problems with continuous variables**

  -[D=1, Gender=1, Age=[20,30)],[D=0, Gender=1, Age=[20,30)]
  
  -[D=1, Gender=0, Age=[30,40)],[D=0, Gender=0, Age=[30,40)]

-- **How to stratify continuous variables**

  -many continuous variables have "natural" breakpoints
  
  -years of education (high school, college, graduate)
  
  -income (official poverty level)
  
  -assumptions: variations between strata and within strata

## Example 2

**This is the idea of matching**

-- Matching as data preprossing to reduce model dependence

<font size="3">
Ho, Daniel E., Kosuke Imai, Gary King, and Elizabeth A. Stuart. "Matching as nonparametric preprocessing for reducing model dependence in parametric causal inference." Political analysis 15, no. 3 (2007): 199-236.
</font>

-- Losses of data and efficiency

-- How to quantify uncertainty

<font size="3">
Iacus, Stefano M., Gary King, and Giuseppe Porro. "A theory of statistical inference for matching methods in causal research." Political Analysis 27, no. 1 (2019): 46-68.
</font>

## Example 2

**Common support and extrapolation**

```{r,warning=FALSE,message=FALSE,echo=FALSE}
ggplot(d)+
  geom_density(aes(Age,fill=as.factor(D)),lwd=1,alpha=0.2)+
  scale_fill_manual(values = c("red","blue"),guide_legend(title="Group"))+
  theme_bw()
```

## Example 2

**Common support and extrapolation**

```{r,warning=FALSE,message=FALSE,echo=FALSE}
ggplot(d)+
  geom_bar(aes(x=factor(D),fill=factor(Gender)),alpha=0.5)+
  scale_fill_manual(values=c("red","blue"),guide_legend(title="Gender"))+
  labs(x="Group",y="")+
  theme_bw()
```

## Example 2
```{r,warning=FALSE,message=FALSE}
### Require the MatchIt package
library(MatchIt)
### Discard data that cannot be matched within strata
mat<-matchit(D~Age+Gender,data=d,method="cem")
mat$nn
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### The new dataset
d_mat<-tibble(get_matches(mat,d));d_mat
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### compare group means within each strata and aggregate (weights)
gpmeans_strata<-lm(data=d_mat,y~D, weights = weight);
summary(gpmeans_strata)$coefficients
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### modelling with the wrong functional form
mod_false<-lm(data=d_mat,y~D+Age+Gender,weights = weight);
summary(mod_false)$coefficients
```

## Example 2

```{r,warning=FALSE,message=FALSE}
### modelling with the right functional form
mod_true<-lm(data=d_mat,y~D+Age+I(Age^2)+Gender,weights = weight);
summary(mod_true)$coefficients
```

## Example 2

**Matching estimators using satuated regressions**

<font size="3">
Angrist, Joshua D., and Jörn-Steffen Pischke. Mostly harmless econometrics: An empiricist's companion. Princeton university press, 2008.
</font>

-- Putting each strata as a dummy

-- No functional forms are assumed

-- A matching estimator with different weights

```{r,warning=FALSE,message=FALSE}
### cut Age to 9 groups with equal size
d<-d %>% mutate(Age_strata=cut_number(Age,n=9))
### Gender, Age_strata and their interactions
mod_mat<-lm(data=d,y~D+Gender*Age_strata)
```

## 

```{r,warning=FALSE,message=FALSE,echo=FALSE}
summary(mod_mat)$coefficients[,1:2]
```

## Take-away Points from Example 2

**In non-experimental studies, we face trade-offs**

<font size="4.5">

-- Modelling helps partion the effect of interests

  - If the functional form is correctly specified
  
  - But we need to know the functional forms

-- There are ways to partion the effect of interest without modelling

  - not model dependent (do not need to assume functional forms)
  
  - The curse of dimensionality
  
  - losses of data and efficiency (structures of the data are not used)
  
  - how to quantify uncertainty

</font>

## Take-away Points from Example 2

**How to deal with the trade-offs**

-- Nonparametric data preprocessing before modelling

-- Present results of both

-- **Data collection: large and informative data**

<font size="4.5">
An example: Chen, Ting, and James Kai-sing Kung. "Busting the “Princelings”: The campaign against corruption in China’s primary land market." The Quarterly Journal of Economics 134, no. 1 (2019): 185-226.
</font>

-- **Research design** (next semester)

## Modelling with Pipes

**Tools that make data exploratory easier (Know your data)**

***Modelr, Purrr, and Broom***

-- *Modelr*: integrating modelling and data with pipes

-- *purrr* and *broom*: seamlessly connect data with many models 

## *Modelr*

```{r,message=FALSE,warning=FALSE}
library(tidyverse);library(modelr)
### read the dataset: media freedom, country-year
d_media<-read_csv("media.csv");d_media
```

## *Modelr*

```{r,message=FALSE,warning=FALSE}
### model matrix
model_matrix(d_media,media_score~gdp_percapita)
```


## *Modelr*

```{r,message=FALSE,warning=FALSE}
### model matrix
model_matrix(d_media,media_score~log(gdp_percapita))
```

## *Modelr*

```{r}
### other transformations
# I (): e.g. I (x^2)
# poly (): polynomial 
# ns (): natural splines from "splines" package
```


## *Modelr*

```{r,message=FALSE,warning=FALSE}
### a simple model
mod_1<-lm(data=d_media,media_score~log(gdp_percapita))
### add predictions to the existing dataset
d_media<-d_media %>% add_predictions(mod_1)
### add residuals to the existing dataset
d_media<-d_media %>% add_residuals(mod_1)
### the basic R function predict() returns a vector
```

## *Modelr*

```{r,message=FALSE,warning=FALSE}
d_media
```

## *Modelr*

```{r,message=FALSE,warning=FALSE,out.width = '70%',fig.align = "center"}
### it's easy to work with
ggplot(d_media,aes(x=log(gdp_percapita)))+
  geom_point(aes(y=media_score))+
  geom_line(aes(y=pred),lwd=1,colour="red")+
  theme_bw()
```

## *Modelr*

```{r,message=FALSE,warning=FALSE,out.width = '70%',fig.align = "center"}
### it's easy to work with
ggplot(d_media,aes(x=log(gdp_percapita)))+
  geom_point(aes(y=resid))+
  theme_bw()
```


## *Modelr*

```{r,message=FALSE,warning=FALSE}
### model matrix: adding another predictor
model_matrix(d_media,media_score~log(gdp_percapita)+log(population))
```

## *Modelr*

```{r,message=FALSE,warning=FALSE}
### a model with both gdp per capita and population
mod_2<-lm(data=d_media,media_score~log(gdp_percapita)+log(population))
```


## *Modelr*

```{r,message=FALSE,warning=FALSE}
### gather residuals/predictions on two (or multiple) models
d_media<-d_media %>% gather_residuals(mod_1,mod_2);d_media
```

## *Modelr*

```{r,message=FALSE,warning=FALSE,out.width = '70%',fig.align = "center"}
ggplot(d_media,aes(x=log(gdp_percapita)))+
  geom_point(aes(y=resid))+
  facet_wrap(~model)+
  theme_bw()
```



## *Modelr*

```{r,message=FALSE,warning=FALSE}
### use grid to customize input data
d_media %>% 
  data_grid(gdp_percapita=seq_range(gdp_percapita,3),
            population=seq_range(population,2)) %>% 
  spread_predictions(mod_1,mod_2)
```

## *Modelr*

```{r,message=FALSE,warning=FALSE}
### use grid to hold other variables constant
d_media %>% 
  data_grid(gdp_percapita=seq_range(gdp_percapita,n=6,trim=0.1),
            .model=mod_2) %>% 
  add_predictions(mod_2)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
library(purrr);library(broom)
### existing dataset: country-year; nest the existing data frame
d_media<-read_csv("media.csv")
d_media_byctry<-d_media %>% 
  group_by(country) %>% 
  nest()
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
d_media_byctry
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
d_media_byctry$data[[1]]
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### fit a model to each country and seamlessly connect with the data
# a function of regression
model_country<-function(df){
  lm(data=df,media_score~log(gdp_percapita))
}
# use map from purrr package to fit a model on each country
d_media_byctry<-d_media_byctry %>% 
  mutate(model=map(data,model_country))
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
d_media_byctry
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### add residuals using map2
d_media_byctry<-d_media_byctry %>% 
  mutate(resids=map2(data,model,add_residuals))
### the order of the argument matters here
# add_residuals(data, model, var = "resid")
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
d_media_byctry
```


## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### use pmap: work with any number of arguments (as a list)
d_media_byctry %>% mutate(resids=pmap(list(data,model),add_residuals))
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### unnest residuals (for plotting or ...)
d_media_byctry %>% unnest(resids) %>% print(n=6)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### broom: turn model outputs into tidy data
tidy(mod_2)
### also work with other hypothesis testing functions 
# tt<-t.test(y~D)
# tidy(tt)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### broom: turn model outputs into tidy data
augment(mod_2)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### broom: use augment to make predictions with confidence intervals
d_media %>% 
  data_grid(gdp_percapita=seq_range(gdp_percapita,6,trim=0.1),
            .model=mod_2) %>% 
  augment(mod_2,newdata=.)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### broom: turn model outputs into tidy data
glance(mod_2)
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
### connect model outputs with the nested data
d_media_byctry<-d_media_byctry %>% 
  mutate(model_out=map(model,glance))
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE}
options(width = 60)
print(d_media_byctry,n=6)
```

## many models with *purrr* and *broom*

Compare the relationship between gdp per capita and media freedom in China, Vietnam, India, and singapore.

```{r,message=FALSE,warning=FALSE}
d_media_byctry<-d_media_byctry %>% 
  mutate(pred_byctry=map(model,augment)) %>% 
  unnest(pred_byctry)
```

## many models with *purrr* and *broom*

Compare the relationship between gdp per capita and media freedom in China, Vietnam, India, and Singapore.

```{r,message=FALSE,warning=FALSE}
print(d_media_byctry,n=5)
```

## many models with *purrr* and *broom*

Compare the relationship between gdp per capita and media freedom in China, Vietnam, India, and Singapore.

```{r,message=FALSE,warning=FALSE}
d_compare<-d_media_byctry %>% 
  filter(country %in% c("China","Vietnam","India","Singapore"))
```

## many models with *purrr* and *broom*

```{r,message=FALSE,warning=FALSE,out.width = '70%',fig.align = "center"}
ggplot(d_compare,aes(x=log.gdp_percapita.))+
  geom_point(aes(y=media_score))+
  geom_line(aes(y=.fitted),lwd=1,colour="red")+
  facet_wrap(~country)+
  theme_bw()
```


